Edge sign prediction in general graphs and Correlation Clustering

### 2017-10-31 edge/model.tex
pass over edge model and methods

We start by introducing some notations and terminology about edge types in node-attributed graphs.
Then we present two options to model the interactions between nodes and the role of attributes in
\enquote{explaining} edges (\autoref{sub:edge_setting}). Choosing of these two options leads us to
formulate the \ecp{} problem in \autoref{sub:edge_constraints}. We then describe additional
constraints to take into account the topology of the graph. Finally we elaborate on the relation
between \ecp{} and \esp{} in signed graphs.

@@

As in the previous chapter, we are given an undirected, unweighted graph $G=(V,E)$. This
time though, we also have side information in addition to the graph topology. Namely, each node $u$
is associated with a $d$-dimensional feature vector $x_u \in \Rbb^d$. As an example, in a social
network, $x_u$ would be the profile of user $u$, describing $u$'s demographics and preferences. We
are also given a combination operator $s$ between feature vectors, such that $s(x_u, x_v) = s_{uv}
\in \Rbb^d$. Finally, we consider distinguished feature vectors $w_{uv}$,
along which the combination $s_{uv}$ between two nodes $u$ and $v$ is evaluated. We call such
$w_{uv}$ \emph{directions}, and constrain them to have unit norm.
While being exactly of norm one is not primordial, the role of bounding the norm of the directions
(\ie{} having $\sum_{i=1}^d {w_{uv}}_{;i}^2 \leq B^2$) is to tie the $d$ dimensions together.
Without such a constraint, we could simply study each dimension separately, whereas this bound
adds a coupling across different attributes, since increasing the weight of an attribute mechanically
affects the others.

@@

This second choice of $s$ and $g$ is indeed more flexible, for it encompasses all cases of
homophily, indifference and heterophily. Furthermore, since the profiles are given, the resulting
$s_{uv}$ can be seen as constants, meaning that $g$ is linear and therefore interpretable. Finally,
it also supports categorical data. We shall demonstrate all these features on a small scale example,
but we first define formally our problem.

@@

In \autoref{fig:edge_exe}, we present a small instance of the \ecp{} problem, where we are given the
graph shown in \autoref{fig:edge_exe_graph}, whose node attributes are listed
\autoref{tab:edge_exe}. The first two attributes are categorical, denoting the company employing
each user, either \textsf{G} or \textsf{F}. Instead of a classic binary one-hot encoding, we
additionally use $-1$ to denote that a user had been employed by a given company. \textsf{Gender} is
encoded as presented earlier, and we also apply a pre processing to transform \textsf{age} between
$-1$ and $1$. Finally, \textsf{opinion} represents the view of former \textsf{F} employees about a
societal issue, with $0$ denoting a neutral position. In \autoref{tab:edge_exe}, we also show two
directions that explain the connections in the graph $G$. The first covers being of the same age and
same companies, especially at \textsf{G}, while the second directions covers former employees of
\textsf{F} having different \textsf{age} and \textsf{opinion}. Not only are the results
interpretable as claimed before, but we can also assess the strength of each explanation, using
their goodness. For instance, we see that the edge between $x$ and $y$ has a low score, because
those two users have indeed the same \textsf{age} and \textsf{opinion}.

@@

In addition to assume that the input graph is the superposition of a small number of layers,
another way to leverage the graph topology is to define local constraints at the node level.
For that we will consider node profiles to be normalized and having unit norm. A practical
justification is that if nodes are
users of a social network and the profiles measure their activity across several domains, this
allows to compare users with various level of total activity. Furthermore, we assume that profiles
are linear combination of directions, that
is $x_u = b_u + \sum_{v \in \nei(u)} a_{uv} w_{uv}$, where $a_{uv}$ are real weights and $b_u
\in \Rbb^d$ is the bias vector of $u$, its inherent profile take in isolation.
Viewed in the other direction, this can be interpreted as follow: each time a node $u$
connect to another node $v$ along a direction $w_{uv}$, it \enquote{consumes} a part of its profile
proportional to $w_{uv}$. Therefore, we subtract the term
\begin{equation}
  \label{eq:edge_node_loss}
  \mathcal{L}_{\mathrm{node}} =
  \sum_{u\in V} \left|\left| x_u - b_u - \sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2_2
\end{equation}
from \eqref{eq:edge_full}.

We also make the assumption that each node $u$ is only involved in $k_\mathrm{local} < k$
different directions. The intuition is that when $d$ is large enough, users have to focus their
energy and can only express their interest in a few number of dimensions. In other words, all the
edges incident to $u$ can only be associated with one of $k_\mathrm{local}$ directions. For
some values of
$k_\mathrm{local}$, this is a \NPc{} graph colouring problem and thus cannot necessarily be enforced
exactly. Still, it can be seen as a learning bias and will be used to generate synthetic data.
Furthermore, we can also express it as an optimization constraint. Specifically, suppose that for
an edge $(u,v)$ associated with the direction $w_i \in \mathcal{D}_k$, we have a $k$-dimensional
vector $y_{uv}$ equal to $e_i$ (that is all zero except for the \ith{} component set to one). Let
us stack all such vectors into a $|E| \times k$ matrix $Y$, and let $C \in \Rbb^{|V| \times |E|}$ be
the incidence matrix of $G$\footnote{The incidence matrix of a graph $G$, usually denoted by $B$, is
the $|V| \times |E|$ matrix such that $B_{i,j} = 1$ if the node $v_i$ and edge $e_j$ are incident
and 0 otherwise.}, with $c_u$ being its $u^{th}$ row. Then $c_u^TY$ is a $k$-dimensional vector that counts
how many times each direction is incident to node $u$. We thus want the $\ell_0$ norm of $c_u^TY$
(\ie the number of its non-zero component) to be upper bounded by $k_\mathrm{local}$.
This is not a convex constraint but we can relax it, for such sparsity inducing problems have been
well studied~\autocite{sparseOptim12}. Namely, we let $y_{uv}$ be a softmax membership
$k$-dimensional vector, that is we set $\forall i \in \rangesk\, y_{uv,i} =
\frac{\exp(\beta{s_{uv}}^T w_i)}{\sum_{j=1}^k \exp(\beta {s_{uv}}^T w_j)} $, with $\beta>0$ a large
positive constant.  And we replace the $\ell_0$ norm by a $\ell_1$ norm (or even a $k$-support
norm~\autocite{KsupportNorm12}), yielding the following $\mathcal{L}_{\mathrm{local}}$ to be
minimized:
\begin{equation}
  \label{eq:edge_local_loss}
  \mathcal{L}_{\mathrm{local}} =
  \sum_{u \in V} \left\| \sum_{v \in \nei(u)} y_{uv} \right\|_1 =
  \sum_{u \in V} \sum_{\ell = 1}^k \sum_{v \in \nei(u)}
  \frac{\exp(\beta{s_{uv}}^T w_\ell)}{\sum_{j=1}^k \exp(\beta {s_{uv}}^T w_j)}
\end{equation}

@@

A simple and natural baseline is to cluster the set of all similarity edge vectors $
\theset{s_{uv}}{(u,v) \in E}$ using the $k$-means algorithm, where $k$ is the number of directions
set in \autoref{p:edge_full}. Formally, given the graph $G=(V,E)$ and the node profiles $X$ as
input, we order the edges from $e_1$ to $e_m$. Then we build the matrix $S \in \Rbb^{m\times d}$, whose
\ith{} row is the similarity $s_{uv} = x_u \circ x_v$ between the profiles of the endpoints of $e_i=(u,v)$.
Running $k$-means on the rows of $S$ will thus partition $E$ into $k$ sets of edges, call them $E_1,
\ldots, E_k$. A drawback of that simplicity is that it does not solve \autoref{p:edge_full}.
Indeed, the partition of $E$ does not provide the set of directions $\mathcal{D}_k = \{w_1, w_2,
\ldots, w_k\} \in \dsphere$ we are looking for. A natural way of obtaining directions from that
partition is to set $w_\ell$ to be the normalized cluster center of $E_\ell$. We refer to this
method as \kmeans{}, and assuming it requires $T$ iterations to converge, it has a linear complexity
of $O(T|E|)$.
\iffalse
\begin{center}
  \rule{\textwidth}{.3pt}
  \begin{algorithmic}[1]
    \Function{\kmeans{}}{graph $G=(V,E=\{e_1, \ldots, e_m \})$, profiles $X$}
    \State build the matrix $S$ such that $S_{i,j} = {s_{e_i}}_{;j}$
    \State let $\yhat{}$ and $\{c_1, \ldots, c_k\}$ be the assignment and centroids resulting of
    running $k$-means on the rows of $S$
    \State \textbf{return} \yhat{}, $\mathcal{D}_k = \left\{\frac{c_\ell}{\left\| c_\ell
    \right\|}\right\}_{\ell=1}^k $
    \EndFunction
  \end{algorithmic}
  \rule{\textwidth}{.3pt}
\end{center}
\fi

However,
this does not guarantee that we have assigned the best direction to every edge. We therefore propose
the following heuristic, inspired by the Lloyd algorithm for $k$-means, but where the directions in
$\mathcal{D}_k$ play the role of centroids. Namely, we alternate between two steps:
\begin{enumerate}[(i), nosep]
  \item create a new partition $\{E'_1, \ldots, E'_k\}$ by assigning every edge $(u,v)$ to the
    direction $w$ maximizing its goodness ${s_{uv}}^T w$, that is $E'_\ell = \theset{(u,v) \in E}{
    w_\ell = \argmax_{w_i \in \mathcal{D}_k} {s_{uv}}^T w_i}$, and
  \item update the directions to make them optimal with respect to the edges currently assigned to
    them. Specifically, set $\mathcal{D}_k = \left\{ w_\ell = \frac{s_\ell}{\left\| s_\ell
    \right\|}\right\}_{\ell=1}^k $, where $s_\ell = \sum_{u,v \in E'_\ell} s_{uv}$ is the sum of all
    the similarity vectors of the edge in $E'_\ell$,
\end{enumerate}
until the edge assignment is stable or we reach a maximum number of iterations. By analogy with the
$k$-means algorithm, obtaining convergence guarantees in the general case would be surprising, but in
practice we expect that only a small number $T$ of iterations would be needed. We call this method
\lloyd{} and note it also has a linear complexity of $O(T|E|)$.

@@

The difficulty in optimizing directly \eqref{eq:edge_full} stems from the inner $\max$ operator, which we
replace here by a convex relaxation. Indeed, one can check that given a set of numbers $S=\{a_1,
a_2, \ldots, a_{|S|}\}$ and real number $\beta > 0$, we have
\begin{equation*}
  \max_{a_i \in S} a_i = \lim_{\beta \rightarrow + \infty} \frac{1}{\beta}
  \log\left( \sum_{i=1}^{|S|} \exp{(\beta a_i)} \right)
\end{equation*}
Using this fact, and averaging the two components of our objective function, we can rewrite
\eqref{eq:edge_full} as:
\begin{equation}
  \label{eq:edge_soft_early}
  \argmax_{{\substack{w_1,\ldots,w_k \in \dsphere,\\
  a_{uv}\in\Rbb,\, b_u \in \Rbb^d}}}
  \frac{1}{|E|\beta} \sum_{u,v\in E}
  \log\Bigl(\sum_{\ell=1}^k\exp\left(\beta {s_{uv}}^T w_\ell \right)\Bigr)
  - \frac{\mu}{|V|}\sum_{u\in V}
  \left|\left| x_u - b_u - \sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2,
\end{equation}
where $\mu > 0$ is a trade-off parameter.
Note that in the second term, because we do not know the edge assignment, we have to decide which of the
$w_1, \ldots, w_k$ is $w_{uv}$ equal to. According to our definition of the goodness function, we
want to let $w_{uv} = \argmax_{w_\ell \in \mathcal{D}_k} {s_{uv}}^T w_\ell$. However, the $\argmax$
function is again not convex and we thus use the same relaxation as before. Namely, this relaxation
$\mathcal{A}$ takes as input a $k$-tuple of directions and return their sum weighted by the softmax
function as follow:
$$ \begin{array}[t]{lrcl}
  \mathcal{A} : & \Rbb^d \times \ldots \times \Rbb^d & \longrightarrow & \Rbb^d \\
      & (w_1,\ldots,w_k) & \longmapsto &
  \sum_{i=1}^k \frac{\exp(\beta{s_{uv}}^T w_i)}{\sum_{j=1}^k \exp(\beta {s_{uv}}^T w_j)} w_i
\end{array} $$
This function has an explicit derivative with respect to any $w_i$, but it is rather costly to
compute and in practice we
found automatic differentiation~\autocite{autograd15} to be more efficient. Whereas the modified
objective function \eqref{eq:edge_soft_early} is now jointly convex in all its variables, the set
$\dsphere$ is not. To ease
the optimization procedure, we thus only impose the directions $w_\ell$ to lie within the unit ball
and expect them to have norm close to one in order to maximize the dot product in the ${s_{uv}}^T
w_\ell$ term. A further simplification is that while \eqref{eq:edge_soft_early} requires to optimize
$kd + |E| + |V|d$ parameters, we fix $a_{uv} = 1$ and $b_u = \onev$ for all edges and nodes in order
to be independent of the size of the graph. The final problem solved by our \combined{} method using
(full) projected gradient descent is therefore:
\begin{equation}
  \label{eq:edge_soft}
\!  \argmax_{w_1,\ldots,w_k \in \dball}
  \frac{1}{|E|\beta} \sum_{u,v\in E}\!
  \log\Bigl(\sum_{\ell=1}^k\exp\left(\beta {s_{uv}}^T w_\ell \right)\Bigr)
  - \frac{\mu}{|V|}\sum_{u\in V}
  \left|\left| x_u  - \sum_{v \in \nei(u)}\!\! \mathcal{A}(w_1,\ldots,w_k) \right|\right|^2
\end{equation}

@@

Besides relaxing the $\max$ operator from \eqref{eq:edge_full}, another solution to remove it would
be to increase $k$ to be equal to $E$. As we already mentioned, this makes the problem trivial. To
avoid this, we add the constraint that all those $|E|$ directions are linear combination of a small
number $k$ of basis directions. In order to formulate this more clearly, we rewrite the previous
objectives with matrices. We first show this can express in a form that is amenable to the
Frank--Wolfe algorithm. Then we take the alternative road of explicit low-rank matrix factorization,
and we conclude by commenting on these approach.

@@

At this stage, we could also add the node loss term \eqref{eq:edge_node_loss}, express in matricial
form as $\|X - B -
C\left(A\circ W^T \right)\|_F^2$, where $B\in \Rbb^{|V| \times d}$ is the node bias vectors $b_u$
stacked vertically, $C \in \Rbb^{|V| \times |E|}$ is the incidence matrix of $G$, $A \in \Rbb^{|E|
\times d}$ is the matrix whose each row is the corresponding $a_{uv}\onev$ vector and $\|\cdot\|_F$
denotes the Frobenius norm. Note that this term is jointly convex in $A$, $B$ and $W$.\footnote{As
one can check that the inequality defining convexity holds for this function.} However, for the
clarity of exposition, we refrain to do so for now.

The remaining issue is that $W\in \mathbb{M}^{d\times |E|}$ is not a convex constraint. As in the
previous formulation, we thus relax it by simply imposing that the $\ell_2$ norm of the columns of
$W$ are not too large. In fact, we already have a bound on these column norms, because of the
nuclear norm constraint.
\begin{prop}
For any $W\in \Rbb^{d\times |E|}$ such that $\|W\|_* \leq \delta$, the \ith{} column $W_{:,i}$ of
$W$ satisfies $\|W_{:,i}\|_2 \leq \delta\sqrt{d}$.
\end{prop}
\begin{proof}
  Let $W$ be a matrix in $\Rbb^{d\times m}$ with $d < m$ such that $\|W\|_* \leq \delta$.
  Furthermore, let its singular value decomposition be such that we can write $W = \sum_{l=1}^d
  \sigma_l u_l v_l^T$, where $u_l$ and $v_l$ are $d$ unit-norm vectors, of dimension $d$ and $m$
  respectively. We can thus express the general term $W_{k,i}$ as $\sum_{l=1}^d \sigma_l v_{l,i}
  u_{l,k}$. Because $v_l$ is unit norm, $v_{l,i} \leq 1$ and $$W_{k,i}^2 \leq \left( \sum_{l=1}^d
  \sigma_l u_{l,k} \right)^2 \leq \sum_{l=1}^d \sigma_l^2 \sum_{l=1}^d u_{l,k}^2$$ using the
  Cauchy--Schwartz inequality. The norm of the \ith{} column of $W$ then satisfies:
  \begin{equation*}
    \|W_{:,i}\|_2 = \left( \sum_{k=1}^m W_{k,i}^2 \right)^\lhalf
    \leq \left( \sum_{k=1}^m \left( \sum_{l=1}^d \sigma_l^2 \sum_{l=1}^d u_{l,k}^2 \right) \right)^\lhalf
    = \left( \left( \sum_{l=1}^d \sigma_l^2 \right) \left( \sum_{l=1}^d \sum_{k=1}^m u_{l,k}^2 \right) \right)^\lhalf
  \end{equation*}
  From $\sum_{l=1}^d |\sigma_l | \leq \delta$, we have that $\sum_{l=1}^d \sigma_l^2 \leq \delta^2$.
  Combined with $u_l$ being unit norm, we conclude that $\|W_{:,i}\|_2 \leq \delta\sqrt{d}$.
\end{proof}
However, setting $\delta = \nicefrac{1}{\sqrt{d}}$ is too stringent in high dimension, and because
this inequality is rather loose, it will likely results in very small columns norm. Instead of
relying solely on the nuclear norm constraint, we therefore add an additional regularization term
equal to $\mu \sum_{i=1}^{|E|} \|W_{:,i}\|_2 = \mu \frop{W}{W}$. Our final convex formulation is
then:
\begin{equation}
  \label{eq:edge_FW}
  \min_{\substack{W\in \Rbb^{d\times |E|} \\ \|W\|_* \leq \delta}} \frop{\mu W - S^T}{W} \,.
\end{equation}

>>>>

We start by introducing some notations and terminology about edge types in node-attributed graphs.
Then we present two options to model the interactions between nodes and the role of attributes in
\enquote{explaining} edges (\autoref{sub:edge_setting}). Choosing one of these two options leads us to
formulate the \ecp{} problem in \autoref{sub:edge_constraints}. We then describe additional
constraints allowing us to take into account the topology of the graph. Finally, we elaborate on the relation
between \ecp{} and \esp{} in signed graphs.

@@

As in the previous chapter, we are given an undirected, unweighted graph $G=(V,E)$. This
time though, we also have side information in addition to the graph topology. Namely, each node $u$
is associated with a $d$-dimensional feature vector $x_u \in \Rbb^d$. As an example, in a social
network, $x_u$ would be the profile of user $u$, describing $u$'s demographics and preferences. We
are also given a combination operator $s$ between feature vectors, such that $s(x_u, x_v) = s_{uv}
\in \Rbb^d$. Finally, we consider distinguished feature vectors $w_{uv}$,
along which the combination $s_{uv}$ between two nodes $u$ and $v$ is evaluated. We call such
$w_{uv}$ \emph{directions}, and constrain them to have unit norm.
While being exactly of norm one is not primordial, the role of bounding the norm of the directions
(\ie{} having $\sum_{i=1}^d {w_{uv}}_{;i}^2 \leq B^2$) is to tie the $d$ dimensions together.
Without such a constraint, we could simply study each dimension separately, whereas this bound
adds a coupling across different attributes, since increasing the weight of an attribute mechanically
affects the weights of the other attributes.

@@

This second choice of $s$ and $g$ is indeed more flexible, for it encompasses all cases of
homophily, indifference and heterophily. Furthermore, since the profiles are given, the resulting
$s_{uv}$ can be seen as constants, meaning that $g$ is linear and therefore interpretable. Finally,
it also supports categorical data. We shall demonstrate all these features on a small example,
but we first define formally our problem.

@@

In \autoref{fig:edge_exe}, we present a small instance of the \ecp{} problem, where we are given the
graph shown in \autoref{fig:edge_exe_graph}, whose node attributes are listed
in Table~\ref{tab:edge_exe}. The first two attributes are categorical, denoting the company employing
each user, either \textsf{G} or \textsf{F}. Instead of a classic binary one-hot encoding, we
additionally use $-1$ to denote that a user had been employed by a given company. \textsf{Gender} is
encoded as presented earlier, and we also apply a pre processing to transform \textsf{age} between
$-1$ and $1$. Finally, \textsf{opinion} represents the view of former \textsf{F} employees about a
societal issue, with $0$ denoting a neutral position. In Table~\ref{tab:edge_exe}, we also show two
directions that explain the connections in the graph $G$. The first covers being of the same age and
same companies, especially at \textsf{G}, while the second directions covers former employees of
\textsf{F} having different \textsf{age} and \textsf{opinion}. Not only are the results
interpretable as claimed before, but we can also assess the strength of each explanation, using
their goodness. For instance, we see that the edge between $x$ and $y$ has a low score, because
although those two users have the same \textsf{age} and used to work at \textsf{F}, they do not
share a position at \textsf{G}.

@@

In addition to assume that the input graph is the superposition of a small number of layers,
another way to leverage the graph topology is to define local constraints at the node level.
For that we will consider node profiles to be normalized and having unit norm. A practical
justification is that if nodes are
users of a social network and the profiles measure their activity across several domains, this
allows to compare users with various level of total activity. Furthermore, we assume that profiles
are linear combination of directions, that
is $x_u = b_u + \sum_{v \in \nei(u)} a_{uv} w_{uv}$, where $a_{uv}$ are real weights and $b_u
\in \Rbb^d$ is the bias vector of $u$, \ie{} its inherent profile taken in isolation.
Viewed in the other direction, this can be interpreted as follow: each time a node $u$
connects to another node $v$ along a direction $w_{uv}$, it \enquote{consumes} a part of its profile
proportional to $w_{uv}$. Therefore, we subtract the term
\begin{equation}
  \label{eq:edge_node_loss}
  \mathcal{L}_{\mathrm{node}} =
  \sum_{u\in V} \left|\left| x_u - b_u - \sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2_2
\end{equation}
from \eqref{eq:edge_full}.

\bigskip

We also make the assumption that each node $u$ is only involved in $k_\mathrm{local} < k$
different directions. The intuition is that when $d$ is large enough, users have to focus their
energy and can only express their interest in a few number of dimensions. In other words, all the
edges incident to $u$ can only be associated with one of $k_\mathrm{local}$ directions. For
some values of
$k_\mathrm{local}$, this is a \NPc{} graph colouring problem and thus cannot necessarily be enforced
exactly. Still, it can be seen as a learning bias and will be used to generate synthetic data.
Furthermore, we can also express it as an optimization constraint. Specifically, suppose that for
an edge $(u,v)$ associated with the direction $w_i \in \mathcal{D}_k$, we have a $k$-dimensional
vector $y_{uv}$ equal to $e_i$ (that is all zero except for the \ith{} component set to one). Let
us stack all such vectors into a $|E| \times k$ matrix $Y$, and let $C \in \Rbb^{|V| \times |E|}$ be
the incidence matrix of $G$\footnote{The incidence matrix of a graph $G$, usually denoted by $B$, is
the $|V| \times |E|$ matrix such that $B_{i,j} = 1$ if the node $v_i$ and edge $e_j$ are incident
and 0 otherwise.}, with $c_u$ being its $u^{th}$ row. Then $c_u^TY$ is a $k$-dimensional vector that counts
how many times each direction is incident to node $u$. We thus want the $\ell_0$ norm of $c_u^TY$
(\ie the number of its non-zero component) to be upper bounded by $k_\mathrm{local}$.
This is not a convex constraint but we can relax it, for such sparsity inducing problems have been
well studied~\autocite{sparseOptim12}. Namely, we let $y_{uv}$ be a softmax membership
$k$-dimensional vector, that is we set $\forall i \in \rangesk\,, y_{uv,i} =
\frac{\exp(\beta{s_{uv}}^T w_i)}{\sum_{j=1}^k \exp(\beta {s_{uv}}^T w_j)} $, with $\beta>0$ a large
positive constant.  And we replace the $\ell_0$ norm by a $\ell_1$ norm (or even a $k$-support
norm~\autocite{KsupportNorm12}), yielding the following $\mathcal{L}_{\mathrm{local}}$ term to be
minimized:
\begin{equation}
  \label{eq:edge_local_loss}
  \mathcal{L}_{\mathrm{local}} =
  \sum_{u \in V} \left\| \sum_{v \in \nei(u)} y_{uv} \right\|_1 =
  \sum_{u \in V} \left( \sum_{i = 1}^k \sum_{v \in \nei(u)}
  \frac{\exp(\beta{s_{uv}}^T w_i)}{\sum_{j=1}^k \exp(\beta {s_{uv}}^T w_j)} \right)
\end{equation}

@@

A simple and natural baseline is to cluster the set of all similarity edge vectors $
\theset{s_{uv}}{(u,v) \in E}$ using the $k$-means algorithm, where $k$ is the number of directions
set in \autoref{p:edge_full}. Formally, given the graph $G=(V,E)$ and the node profiles $X$ as
input, we order the edges from $e_1$ to $e_m$. Then we build the matrix $S \in \Rbb^{m\times d}$, whose
\ith{} row is the similarity $s_{uv} = x_u \circ x_v$ between the profiles of the endpoints of $e_i=(u,v)$.
Running $k$-means on the rows of $S$ will thus partition $E$ into $k$ sets of edges, call them $E_1,
\ldots, E_k$. A drawback of that simplicity is that it does not solve \autoref{p:edge_full}.
Indeed, the partition of $E$ does not provide the set of directions $\mathcal{D}_k = \{w_1, w_2,
\ldots, w_k\} \in \dsphere$ we are looking for. A natural way of obtaining directions from that
partition is to set $w_\ell$ to be the normalized cluster center of $E_\ell$. We refer to this
method as \kmeans{}, and assuming it requires $T$ iterations to converge, it has a linear complexity
of $O(Td|E|)$.
\iffalse
\begin{center}
  \rule{\textwidth}{.3pt}
  \begin{algorithmic}[1]
    \Function{\kmeans{}}{graph $G=(V,E=\{e_1, \ldots, e_m \})$, profiles $X$}
    \State build the matrix $S$ such that $S_{i,j} = {s_{e_i}}_{;j}$
    \State let $\yhat{}$ and $\{c_1, \ldots, c_k\}$ be the assignment and centroids resulting of
    running $k$-means on the rows of $S$
    \State \textbf{return} \yhat{}, $\mathcal{D}_k = \left\{\frac{c_\ell}{\left\| c_\ell
    \right\|}\right\}_{\ell=1}^k $
    \EndFunction
  \end{algorithmic}
  \rule{\textwidth}{.3pt}
\end{center}
\fi

However,
this does not guarantee that we have assigned the best direction to every edge. We therefore propose
the following heuristic, inspired by the Lloyd algorithm for $k$-means, but where the directions in
$\mathcal{D}_k$ play the role of centroids. Namely, we alternate between two steps:
\begin{enumerate}[(i), nosep]
  \item create a new partition $\{E'_1, \ldots, E'_k\}$ by assigning every edge $(u,v)$ to the
    direction $w$ maximizing its goodness ${s_{uv}}^T w$, that is $E'_\ell = \theset{(u,v) \in E}{
    w_\ell = \argmax_{w_i \in \mathcal{D}_k} {s_{uv}}^T w_i}$, and
  \item update the directions to make them optimal with respect to the edges currently assigned to
    them. Specifically, set $\mathcal{D}_k = \left\{ w_\ell = \frac{s_\ell}{\left\| s_\ell
    \right\|}\right\}_{\ell=1}^k $, where $s_\ell = \sum_{u,v \in E'_\ell} s_{uv}$ is the sum of all
    the similarity vectors of the edge in $E'_\ell$,
\end{enumerate}
until the edge assignment is stable or we reach a maximum number of iterations. By analogy with the
$k$-means algorithm, obtaining convergence guarantees in the general case would be surprising, but in
practice we expect that only a small number $T$ of iterations would be needed. We call this method
\lloyd{} and note it also has a linear complexity of $O(Td|E|)$.

@@

The difficulty in optimizing directly \eqref{eq:edge_full} stems from the inner $\max$ operator, which we
replace here by a convex relaxation. Indeed, one can check that given a set of numbers $S=\{a_1,
a_2, \ldots, a_{|S|}\}$ and real number $\beta > 0$, we have
\begin{equation*}
  \max_{a_i \in S} a_i = \lim_{\beta \rightarrow + \infty} \frac{1}{\beta}
  \log\left( \sum_{i=1}^{|S|} \exp{(\beta a_i)} \right)
\end{equation*}
Using this fact, and averaging the two components of our objective function, we can rewrite
\eqref{eq:edge_full} as:
\begin{equation}
  \label{eq:edge_soft_early}
  \argmax_{{\substack{w_1,\ldots,w_k \in \dsphere,\\
  a_{uv}\in\Rbb,\, b_u \in \Rbb^d}}}
  \frac{1}{|E|\beta} \sum_{u,v\in E}
  \log\Bigl(\sum_{\ell=1}^k\exp\left(\beta {s_{uv}}^T w_\ell \right)\Bigr)
  - \frac{\mu}{|V|}\sum_{u\in V}
  \left|\left| x_u - b_u - \sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2,
\end{equation}
where $\mu > 0$ is a trade-off parameter.
Note that in the second term, because we do not know the edge assignment, we have to decide which of the
$w_1, \ldots, w_k$ is $w_{uv}$ equal to. According to our definition of the goodness function, we
want to let $w_{uv} = \argmax_{w_\ell \in \mathcal{D}_k} {s_{uv}}^T w_\ell$. However, the $\argmax$
function is again not convex and we thus use the same relaxation as before. Namely, this relaxation
$\mathcal{A}$ takes as input a $k$-tuple of directions and return their sum weighted by the softmax
function as follow:
$$ \begin{array}[t]{lrcl}
  \mathcal{A} : & \Rbb^d \times \ldots \times \Rbb^d & \longrightarrow & \Rbb^d \\
      & (w_1,\ldots,w_k) & \longmapsto &
  \sum_{i=1}^k \frac{\exp(\beta{s_{uv}}^T w_i)}{\sum_{j=1}^k \exp(\beta {s_{uv}}^T w_j)} w_i
\end{array} $$
This function has an explicit derivative with respect to any $w_i$, but it is rather costly to
compute and in practice we
found automatic differentiation~\autocite{autograd15} to be more efficient. Whereas the modified
objective function \eqref{eq:edge_soft_early} is now jointly convex in all its variables, the set
$\dsphere$ is not. To ease
the optimization procedure, we thus only impose the directions $w_\ell$ to lie within the unit ball
and expect them to have norm close to one in order to maximize the dot product in the ${s_{uv}}^T
w_\ell$ term\footnote{Indeed, since the first term is convex, its maximum can only be reach at the
boundary of the feasible domain~\autocite[Theorem 32.1]{convexAnalysis70}. However, we are also
subtracting another convex term and the resulting function is therefore more complicated.}.
A further simplification is that while \eqref{eq:edge_soft_early} requires to optimize
$kd + |E| + |V|d$ parameters, we fix $a_{uv} = 1$ and $b_u = \onev$ for all edges and nodes in order
to be independent of the size of the graph. The final problem solved by our \combined{} method using
(full) projected gradient descent is therefore:
\begin{equation}
  \label{eq:edge_soft}
\!  \argmax_{w_1,\ldots,w_k \in \dball}
  \frac{1}{|E|\beta} \sum_{u,v\in E}\!
  \log\Bigl(\sum_{\ell=1}^k\exp\left(\beta {s_{uv}}^T w_\ell \right)\Bigr)
  - \frac{\mu}{|V|}\sum_{u\in V}
  \left|\left| x_u  - \sum_{v \in \nei(u)}\!\! \mathcal{A}(w_1,\ldots,w_k) \right|\right|^2
\end{equation}

@@

Besides relaxing the $\max$ operator from \eqref{eq:edge_full}, another solution to remove it would
be to increase $k$ to be equal to $E$. As we already mentioned, this makes the problem trivial. To
avoid this, we add the constraint that all those $|E|$ directions are linear combinations of a small
number $k$ of basis directions. In order to formulate this more clearly, we rewrite the previous
objectives with matrices. We first show this can be expressed in a form that is amenable to the
Frank--Wolfe algorithm. Then we take the alternative road of explicit low-rank matrix factorization,
and we conclude by commenting on these approaches.

@@

At this stage, we could also add the node loss term \eqref{eq:edge_node_loss}, expressed in matricial
form as $\|X - B -
C\left(A\circ W^T \right)\|_F^2$, where $B\in \Rbb^{|V| \times d}$ is the node bias vectors $b_u$
stacked vertically, $C \in \Rbb^{|V| \times |E|}$ is the incidence matrix of $G$, $A \in \Rbb^{|E|
\times d}$ is the matrix whose each row is the corresponding $a_{uv}\onev$ vector and $\|\cdot\|_F$
denotes the Frobenius norm. Note that this term is jointly convex in $A$, $B$ and $W$.\footnote{As
one can check that the inequality defining convexity holds for this function.} However, for the
clarity of exposition, we refrain to do so for now.

The remaining issue is that $W\in \mathbb{M}^{d\times |E|}$ is not a convex constraint. As in the
previous formulation, we thus relax it by simply imposing that the $\ell_2$ norm of the columns of
$W$ are not too large. In fact, we already have a bound on these column norms, because of the
nuclear norm constraint.
\begin{prop}
For any $W\in \Rbb^{d\times |E|}$ such that $\|W\|_* \leq \delta$, the \ith{} column $W_{:,i}$ of
$W$ satisfies $\|W_{:,i}\|_2 \leq \delta\sqrt{d}$.
\end{prop}
\begin{proof}
  Let $W$ be a matrix in $\Rbb^{d\times m}$ with $d < m$ such that $\|W\|_* \leq \delta$.
  Furthermore, let its singular value decomposition be such that we can write $W = \sum_{l=1}^d
  \sigma_l u_l v_l^T$, where $u_l$ and $v_l$ are $d$ unit-norm vectors, of dimension $d$ and $m$
  respectively. We can thus express the general term $W_{k,i}$ as $\sum_{l=1}^d \sigma_l v_{l,i}
  u_{l,k}$. Because $v_l$ is unit norm, $v_{l,i} \leq 1$ and $$W_{k,i}^2 \leq \left( \sum_{l=1}^d
  \sigma_l u_{l,k} \right)^2 \leq \sum_{l=1}^d \sigma_l^2 \sum_{l=1}^d u_{l,k}^2$$ using the
  Cauchy--Schwartz inequality. The norm of the \ith{} column of $W$ then satisfies:
  \begin{equation*}
    \|W_{:,i}\|_2 = \left( \sum_{k=1}^m W_{k,i}^2 \right)^\lhalf
    \leq \left( \sum_{k=1}^m \left( \sum_{l=1}^d \sigma_l^2 \sum_{l=1}^d u_{l,k}^2 \right) \right)^\lhalf
    = \left( \left( \sum_{l=1}^d \sigma_l^2 \right) \left( \sum_{l=1}^d \sum_{k=1}^m u_{l,k}^2 \right) \right)^\lhalf
  \end{equation*}
  From $\sum_{l=1}^d |\sigma_l | \leq \delta$, we have that $\sum_{l=1}^d \sigma_l^2 \leq \delta^2$.
  Combined with $u_l$ being unit norm, we conclude that $\|W_{:,i}\|_2 \leq \delta\sqrt{d}$.
\end{proof}
However, setting $\delta = \nicefrac{1}{\sqrt{d}}$ is too stringent in high dimension, and because
this inequality is rather loose, it will likely results in very small column norms. Instead of
relying solely on the nuclear norm constraint, we therefore add an additional regularization term
equal to $\mu \sum_{i=1}^{|E|} \|W_{:,i}\|_2 = \mu \frop{W}{W}$. Our final convex formulation is
then:
\begin{equation}
  \label{eq:edge_FW}
  \min_{\substack{W\in \Rbb^{d\times |E|} \\ \|W\|_* \leq \delta}} \frop{\mu W - S^T}{W} \,.
\end{equation}
